Title

### 2016-05-24 paper/sections/Overview.tex
Revision on Sec 2 & 3

\section{Overview}
\weixin{title: Evolution on DSLs?}

@@

%As language evolves, the need of new syntax often arises.
As DSLs evolve along the time, the demand for new syntax and and semantics may arise.
It would be good if we can introduce these new features to the DSL \emph{modularly}.
%This is actually a hard problem, known as the Expression Problem (EP)~\cite{wadler}, which
This, however, requires the host language equipped with two dimensions of
extensibility.
In this section, we argue that from extensibility perspective, OO languages are
better candidates as host languages than FP languages.
To demonstrate this, we try to extend \dsl with new syntax and new semantics in
the setting of shallow embedding. Adding new syntax is both easy for both FP
and OOP. Adding new semantics, however, is hard in FP. Although it is possible
in OOP, the solution requires some boilerplate code. We hence developed \name
for defining modular extensions easily.

\subsection{Initial System}
Before introducing any extensions, we need to rewrite the initial implementation for \dsl shown in Figure~\ref{}:

\lstinputlisting[linerange=11-26]{../src/paper/sec3/Circuit.java}%APPLY:INIT
We change the implementation language constructs from classes to interfaces and
their fields to unimplemented getter methods. The purpose of this modification
is to allow types of the fields to be refined in future, which is vital for
retaining extensibility.

@@

\lstinputlisting[linerange=29-40]{../src/paper/sec3/Circuit.java}%APPLY:SYNTAX

@@

Adding new semantics, however, becomes hard for shallow embedding.
Gibbons and Wu worked around this problem in the following way:
\lstinputlisting[language=haskell,linerange=2-6]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
which is similar to how we add \texttt{desugar} to the implementation in Section~\ref{}.
The difference is that they use a tuple instead of a record to merge semantic
domains, then define the two semantic functions simultaneously for each
case, and split the semantic functions through projections on the tuple in the end.
This solution, however, modifies existing code, breaking the requirement of EP.

@@

\name addresses all these problems through embracing family polymorphism and code instrument.
By using \name, we can refactor the extensions in the way shown in Figure\ref{}.
\lstinputlisting[linerange=80-82]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY
\lstinputlisting[linerange=110-112]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY_SYNTAX
\lstinputlisting[linerange=131-156]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY_SEMANTICS

@@

\lstinputlisting[linerange=164-176]{../src/paper/sec3/Circuit.java}%APPLY:INSTRUMENT
% TODO: not implemented yet
Family polymorphism on only gives more safety on the client code but also helps
language implementers catch bug early.
For example, if one forgets to implement, say cases from \texttt{Family}, in
\texttt{Semantics}, she would get warned they have not implemented the
\texttt{wellSized} method because class-like interfaces should not contain any
unimplemented methods except for getters.
These errors are captured by automatically re-declaring members with inheritance relationships.
%The \texttt{of} method is  returns an instance of an anonymous class
%that implements the interface with all getters implemented and its instance is returned.

% Multiple interpretation
% Dependent interpretation
% Context-sensitive interpretation

>>>>

\section{Evolution on DSLs with \name}

@@

%As language evolves, the need of new syntax often arises.
As DSLs evolve along the time, the demand for new syntax and new semantics may arise.
It would be good if we can introduce these new features to the DSL \emph{modularly}.
%This is actually a hard problem, known as the Expression Problem (EP)~\cite{wadler}, which
%This, however, requires the host language equipped with two dimensions of extensibility.
For shallow embedding, adding new syntax is trivial - just adding new cases in
the semantic function. On the other hand, adding new semantics becomes hard as it allows
only one definition for each case. Gibbons and Wu summarized how to define
various type of semantic functions (interpretations) in shallow embedding by
applying \emph{fold theory}. However, the encoding is based on tuples, which is
not modular extensible. Though they also presented modular solution based on
datatype a lar carte\cite{}. But such solution changes the encoding style
dramatically and may be too complex to use practically.
In contrast, the OO approach allows all kinds of
interpretations to be defined in an modular intuitive way.
In this section, we will show how various types of interpretations can be
defined using the OO way with the help of \name. Along the way,
some features of \name will be introduced, which simplify our implementation significantly.

%argue that from extensibility perspective, OO languages are
%better candidates as host languages than FP languages.
%To demonstrate this, we try to extend \dsl with new syntax and new semantics in
%the setting of shallow embedding. Adding new syntax is both easy for both FP
%and OOP. Adding new semantics, however, is hard in FP. Although it is possible
%in OOP, the solution requires some boilerplate code. We hence developed \name
%for defining modular extensions easily.

\subsection{Initial System}
Recall the \dsl language discussed in Section~\ref{}.
We redefine it with \name:

\lstinputlisting[linerange=19-35]{../src/paper/sec3/Circuit.java}%APPLY:INIT

There are three main changes:
1) All the original implementations are moved inside an interface \texttt{Init}; and
2) The constructor classes are now interfaces for the purpose of employing
multiple interface inheritance of Java;
3) Fields of the classes become getter methods (starting with an underscore) so that their types can be refined for retaining extensibility~\cite{}.

% introduce of method
As classes are now interfaces, how can we create objects on them?
\name automatically generates a \emph{static} \texttt{of} method for each
class-like interface, i.e. an interface that contains only getters, for creating objects.
For instance, the \texttt{of} method for \texttt{Beside} is:
\begin{lstlisting}
static Beside of(Circuit c1, Circuit c2) {
  return new Beside() {
    Circuit _c1 = c1; Circuit _c2 = c2;
    public Circuit _c1() { return c1; }
    public Circuit _c2() { return c2; }
  }
}
\end{lstlisting}
Inside the \texttt{of} method, an anonymous class that implements
\texttt{Beside} is created and its instance is returned.

\subsection{Multiple Interpretations}
Defining additional interpretations is not so trivial for shallow embedding,
especially for functional languages.
Suppose that we want to have a function that checks whether a circuit is constructed correctly.

Here comes the solution from Gibbons and Wu.
\lstinputlisting[language=haskell,linerange=2-8]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
where tuples are used to aggregate multiple interpretations in one definition.
The way to introduce \texttt{desugar} in previous section is similar to this
approach, where we use records instead.
However, neither tuples nor records are extensible - whenever we need a new
interpretation, we have to revise the original code, either expanding the arith of the
tuple or inserting a new field in the record.

Conversely, the OO approach allows us to define multiple interpretations in a
\emph{modular} way through extending existing code:
\lstinputlisting[linerange=39-52]{../src/paper/sec3/Circuit.java}%APPLY:MULTIPLE
We expand the semantics in an extended family \texttt{Multiple}.
Inside \texttt{Multiple}, the \texttt{Circuit} is re-declared
with a new method \texttt{wellSized} and the \texttt{Circuit} hierarchy
implements the \texttt{wellSized} method.
Through extending the \texttt{Init} family, the \texttt{width} method is
inherited without duplicating its definition. This way we modularly extend the
semantics.

Note that we do not explicitly write down \texttt{extends} clauses in the
extended family.
Instead, they are inferred by \name according to their
names and the inheritance relationship between families.
After instrumentation done by \name, the definitions of \texttt{Circuit} and \texttt{Beside}
would be:

\begin{lstlisting}
interface Circuit extends Init.Circuit {
  ... // wellSized declaration
}
interface Beside extends Circuit, Init.Beside {
  Circuit _c1(); Circuit _c2(); // type refinements
  ... // wellSized implementation
}
\end{lstlisting}
Interfaces from base families of the same name will be put after \texttt{extends} clause and the
interface hierarchy of \texttt{Circuit} is preserved in the extended family.
Moreover, \name re-declare getters from the base \texttt{Beside} in the extended
\texttt{Beside} so that their return types are refined to be the new
\texttt{Circuit}.

\subsection{Dependent Interpretations}
Interpretations which depend on other interpretations are called \emph{dependent
  interpretation}. %They are hard to express modularly in traditional shallow embedding.
Before we introduce an example of such interpretations, we first add two new
constructs to our language:
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[language=haskell,linerange=40-41]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

%\lstinputlisting[linerange=56-69]{../src/paper/sec3/Circuit.java}%APPLY:SYNTAX
%With these new constructs, more complex circuits can be constructed.
%For example, Figure~\ref{} shows the circuit constructed by \emph{stretch
%  [2,2,2] (fan 3) `beside` fan 1}.

To add these two new constructs to \dsl, we can declare two class-like interfaces \texttt{Above} and
\texttt{Stretch} that extends \texttt{Circuit} with \texttt{width} and
\texttt{wellSized} implemented

\lstinputlisting[linerange=56-76]{../src/paper/sec3/Circuit.java}%APPLY:DEPENDENT
Definitions of \texttt{wellSized} for the new cases make it a dependent interpretation, as they
rely on the \texttt{width} method for calculating the width of a circuit.

%Since \texttt{Circuit} in \texttt{Dependent} family extends the one from
%\texttt{Init}, we can simply call \texttt{width} method on a
%\texttt{Circuit} object when the width of a circuit is needed in the definition
%of \texttt{wellSized}.

\subsection{Context-sensitive Interpretations}
% add tlayout
Connections between vertical wires of a circuit can be modeled by dividing the
circuit into different layers, with local connections only go rightwards from
one layer to the next. Each layer is a collection of pairs $(i,j)$ denoting a
connection from wire $i$ on this layer to wire $j$ on the next. The circuit shown in
Figure\ref{} has the following connections:

$$[[(0,1),(2,3)],[(1,3)],[(1,2)]]$$

The above shows that the circuit has 3 layers: the first layer has connections
from the first wire to the second, and from the second to the third; the second
layer has only one connection from the first wire to the third wire; while the
third layer has a connection from the first wire to the second.

The method \texttt{tlayout} deduce the connections of a circuit. It additionally
takes a parameter \texttt{}.

\lstinputlisting[linerange=98-130]{../src/paper/sec3/Circuit.java}%APPLY:CONTEXT_SENSITIVE

\texttt{tlayout} takes a transformation over \texttt{IntUnaryOperator} is considered as the context which might be changed.
The accumulating argument \texttt{f} is a transformation function over the indices of wires in a circuit.

Some definitions of helper methods are omitted for space reason.
Their functionalities are: \texttt{lzw} stands for 'long zip with', which takes two lists and a binary operator for
combining elements on the same position; if one list is shorter then the
remaining elements of the other are appended to the result list; \texttt{concat}
concatenates two lists with the first tail connected the second head;
The indices of wires are calculated by accumulating the sum of the list \texttt{\_ns} using the \texttt{scanl1}.

% in \texttt{CtxSensitive}

\subsection{Interpretation Combination}
One appealing feature that \name approach provides is that we can modularly
merge interpretations defined for the same data structure together.
This is done through from different families
together using multiple interface inheritance.

For example, we can merge the interpretations we have defined so far in this way:

\lstinputlisting[linerange=161-163]{../src/paper/sec3/Circuit.java}%APPLY:MERGE
Simply defining a new family \texttt{Merge} that extends other families we want
to combine. \name will implicitly re-declare all the interfaces and inject the
inheritance relationships. For example, the implicit redeclaration of
\texttt{Circuit} looks like this:

\begin{lstlisting}
interface Circuit extends Multiple.Circuit, Dependent.Circuit, CtxSensitive.Circuit {}
\end{lstlisting}

So do other interfaces.

\begin{comment}
\subsection{Parametrized interpretations}
\subsection{Implicitly Parametrized interpretations}
\subsection{Modular Interpretations}

@@

\lstinputlisting[linerange=56-69]{../src/paper/sec3/Circuit.java}%APPLY:SYNTAX

@@

Adding new semantics, however, becomes hard for shallow embedding.
Gibbons and Wu worked around this problem in the following way:
\lstinputlisting[language=haskell,linerange=2-8]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
which is similar to how we add \texttt{desugar} to the implementation in Section~\ref{}.
The difference is that they use a tuple instead of a record to merge semantic
domains, then define the two semantic functions simultaneously for each
case, and split the semantic functions through projections on the tuple in the end.
This solution, however, modifies existing code, breaking the requirement of EP.

@@

\name addresses all these problems through embracing family polymorphism and code instrument.
By using \name, we can refactor the extensions in the way shown in Figure\ref{}.
\lstinputlisting[linerange=46-48]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY
\lstinputlisting[linerange=76-78]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY_SYNTAX
\lstinputlisting[linerange=97-122]{../src/paper/sec3/Circuit.java}%APPLY:FAMILY_SEMANTICS

@@

\lstinputlisting[linerange=168-180]{../src/paper/sec3/Circuit.java}%APPLY:INSTRUMENT
% TODO: not implemented yet
Family polymorphism on only gives more safety on the client code but also helps
language implementers catch bug early.
For example, if one forgets to implement, say cases from \texttt{Family}, in
\texttt{Semantics}, she would get warned they have not implemented the
\texttt{wellSized} method because class-like interfaces should not contain any
unimplemented methods except for getters.
These errors are captured by automatically re-declaring members with inheritance relationships.
%The \texttt{of} method is  returns an instance of an anonymous class
%that implements the interface with all getters implemented and its instance is returned.
\end{comment}

### 2016-05-24 paper/sections/ShallowOO.tex
Revision on Sec 2 & 3

Weixin writes this part.

@@

\lstinputlisting[language=haskell,linerange=2-5]{./code/shallow.hs}%APPLY:EXPR1
\lstinputlisting[language=haskell,linerange=9-12]{./code/shallow.hs}%APPLY:EXPR2
\lstinputlisting[language=haskell,linerange=16-18]{./code/shallow.hs}%APPLY:EXPR3
\lstinputlisting[language=haskell,linerange=22-33]{./code/shallow.hs}%APPLY:EXPR4

>>>>

\begin{comment}
Weixin writes this part.

@@

Shallow embedding interpreter pattern
Deep embedding visitor pattern
\end{comment}

In this section, we argue that \emph{shallow embedding} and straightforward OO
programming are closely related. To demonstrate it, we use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{} as the running example.
We first give the original implementation of the language in shallow embedding
and then gradually rewrite the definition into an ``OO style''. Also, we show how to do
transformations in shallow embedding, which is typically thought to be the
strength of deep embedding.

\subsection{A DSL for Parallel Prefix Circuits}
Consider an DSL named \dsl that models parallel circuits.
Its BNF grammar is given below:
$$
\begin{array}{lcl}
Circuit & ::= & Identity \\
 & \mid & Fan \\
 & \mid & Beside \\
Fan & ::= & ``fan''\; PositiveNumber \\
Identity & ::= & ``identity''\;  PositiveNumber \\
Beside & ::= & Circuit ``beside'' Circuit \\
PositiveNumber & ::= & [1-9] \\
     & \mid & [1-9][0-9]^* \\
\end{array}
$$
% bnf
\weixin{the bnf grammar needs polishing}

% BNF grammar should not contain left recursion and left factor, rewrite it
% using descent recursion

There are three constructs - two primitives
\emph{identity} and \emph{fan} and one combinator \emph{beside}, whose
meanings are: \emph{identity n} contains \emph{n} isolated vertical wires;
\emph{fan n} contains \emph{n} vertical wires and the first wire connects to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
For example, the circuit \emph{fan 3 beside identity 3} can be visualized in Figure~\ref{}.

\subsection{Shallow Embedding and Object-Oriented Programming}
Different from deep embedding, which uses a data structure to represent the
abstract syntax of a language, shallow embedding defines a language
directly through encoding its semantics. In the case of \dsl,
the shallow embedding implementation should conform to the following types:

\lstinputlisting[language=haskell,linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[language=haskell,linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \texttt{Circuit}, representing the semantic domain, is not defined yet
and should be filled in later with a concrete type according to the semantics.
For example, if the semantics of \dsl is to calculate the width of a
circuit, the semantic function would be:
\lstinputlisting[language=haskell,linerange=11-16]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
As shallow embedding is semantics-oriented, it is already an integer after
constructing a circuit and hence \texttt{width1} is an identity function.
For example, we will immediately get $6$ for
\texttt{fan 3 `beside` identity 3} and hence \texttt{width1} is just an identity function.

The above implementation can be rewritten by wrapping the result into an
datatype and get back the value through pattern matching:
\lstinputlisting[language=haskell,linerange=20-26]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
Then circuit construction gives back a \texttt{Circuit2} which contains an
integer value representing the width and \texttt{width2} extracts out the value.
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

The rewrite goes on by using a record with a sole field to represent the result:
\lstinputlisting[language=haskell,linerange=31-36]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
We no longer need a separate \texttt{width3} function as the field name is the extractor.
The implementation is still shallow since essentially it is nothing different
from the last one.
However, the definition already approaches an OO program such that we can easily
translate it into an OO language like Java:
\lstinputlisting[linerange=7-26]{../src/paper/sec2/Circuit.java}%APPLY:CIRCUIT_OO
where the datatype maps to the abstract class \texttt{Circuit} with its field declared as an abstract method.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \texttt{Circuit} by giving \texttt{width} a definition.

This implementation is exactly how we model \dsl in an OO language for the first
place, following the \interp pattern.
In other words, shallow embedding and straightforward OO programming are closely
related (deep embedding is closely related to the \textsc{Visitor} pattern).

\subsection{Transformations in Shallow Embedding}
One of the main criticism of shallow embedding is that how to define transformations
becomes unclear since there is not a data structure representing the Abstract Syntax Tree (AST).
%In fact, transformations can still be defined in shallow embedding.

An instance of this kind of operations could be desugaring, which eliminates some
language constructs that could be expressed by other language constructs without
sacrificing the expressiveness of the language.
In \dsl, the \emph{identity n} construct is such a syntactic sugar that can
be rewritten via the following formula:
$$identity\ n = (fan\ 1) \overbrace{\ beside\ (fan\ 1)}^{n-1}\ (n > 0)$$
which states that \emph{identity n} can be represented as \emph{n} of \emph{fan 1} combined with \emph{beside}.
%Using the above formula, we can define a transformation that desugars every
%occurrence of \emph{identity} to a combination of \emph{fan} and \emph{beside}.

Implementing the desugaring method in Java could be:
\lstinputlisting[linerange=31-61]{../src/paper/sec2/Circuit.java}%APPLY:DESUGAR_OO
The newly introduced method \texttt{desugar} in \texttt{Circuit} returns a \texttt{Circuit} reflecting that it is a transformation. The implementation of \texttt{desugar} for each case is
straightforward: \emph{fan n} just returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \texttt{Beside}; \emph{identity n} mimics the
formula.

\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Java implementation, we can also define \texttt{desugar} in Haskell:

\lstinputlisting[language=haskell,linerange=6-21]{./code/Desugar.hs}%APPLY:DESUGAR
The definition relies on the laziness of Haskell.
